Statistical Methods

### 2022-07-10 notes/theory.tex
more note updates

\renewcommand{\P}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\V}{\text{V}}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\cov}{cov}

@@

\begin{align}
  \log\mathcal{L} = \sum_{v \in \mathcal{V}} \sum_{i \ne j \in \mathcal{S}} \log(P(O_{i,j}(v) | \theta))
\end{align}

@@

\begin{align}
  \log\mathcal{L} &= \sum_{v \in \mathcal{V}} \left[\log(\pi(v | \theta)) n_\text{D}(v) + \log(1-\pi(v | \theta)) n_\text{S}(v)\right] \\
                  &= \sum_{v \in \mathcal{P}} \left[\log(\pi(v | \theta)) n_\text{D}(v) + \log(1-\pi(v | \theta)) n_\text{S}(v)\right] + \sum_{v \in \mathcal{F}} \log(1-\pi(v | \theta)) n_T  \\
\end{align}

>>>>

\renewcommand{\P}{\mathbb{P}}
\renewcommand{\vec}[1]{\mathbf{#1}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\V}{\text{V}}
\DeclareMathOperator{\var}{Var}
\DeclareMathOperator{\cov}{cov}

@@

\begin{align}
  \log\mathcal{L}(\theta) = \sum_{v \in \mathcal{V}} \sum_{i \ne j \in \mathcal{S}} \log(P(O_{i,j}(v) | \theta))
\end{align}

@@

\begin{align}
  \log\mathcal{L}(\theta) &= \sum_{v \in \mathcal{V}} \left[\log(\pi(v | \theta)) n_\text{D}(v) + \log(1-\pi(v | \theta)) n_\text{S}(v)\right] \\
                  &= \sum_{v \in \mathcal{P}} \left[\log(\pi(v | \theta)) n_\text{D}(v) + \log(1-\pi(v | \theta)) n_\text{S}(v)\right] + \sum_{v \in \mathcal{F}} \log(1-\pi(v | \theta)) n_T(v)  \\
\end{align}

In practice, we calculate these values across bins. For each bin, we treat
$\pi(v | \theta)$ as fixed, assuming that at this scale, the variation in
expected diversity across sites is minimal. For a particular chromosome, we
have two classes of sites: those included in the diversity calculation and
those ignored. The former sites are all putatively neutral and have reliably
called genotypes, and the other sites are possibly non-neutral or do have
reliably called genotypes. The log-likelihood is the sum of bin
likelihoods, $\mathcal{L}(b)$

\begin{align}
  \log\mathcal{L}(b | \theta)  &= \log(\bar{\pi}(b | \theta)) \sum_{v \in \mathcal{P}} n_\text{D}(v) + \log(1-\bar{\pi}(b | \theta)) \left(\sum_{v \in \mathcal{V}} n_\text{S}(v) +  \sum_{v \in \mathcal{F}} n_T(v)  \right).
\end{align}

We can assume that the total number of combinations at each fixed site is
constant, e.g. $n_T = n_T(v)$ for all $v$, simplifying this as

\begin{align}
  \log\mathcal{L}(b | \theta)  &= \log(\bar{\pi}(b | \theta)) \sum_{v \in \mathcal{P}} n_\text{D}(v) + \log(1-\bar{\pi}(b | \theta)) \left(\sum_{v \in \mathcal{V}} n_\text{S}(v) +  n_T(v) f(b)  \right)
\end{align}

where $f(b)$ is the number of fixed sites in bin $b$. Thus, we our data can be
reduced within a bin by these terms summed over sites in the window,

\begin{align}
  \log\mathcal{L}(b | \theta)  &= \log(\bar{\pi}(b | \theta)) Y_D(b) + \log(1-\bar{\pi}(b | \theta)) Y_S(b)
\end{align}

\begin{align}
  \log\mathcal{L}(b | \theta)  &= \log(\bar{\pi}_0 B(b | \vec{w}, \vec{t})) Y_D(b) + \log(1-\bar{\pi}_0 B(b |  \vec{w}, \vec{t})) Y_S(b)
\end{align}

where $\vec{w}$ and $\vec{t}$ are the vectors of mutation rates into

There are two approaches: interpolated $\pi(m | \theta)$ at the midpoint of the
bin position $m$, and averaged bin-averaged $\bar{\pi}(m | \theta)$.
